Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-03-06 rmd/manuscript.Rmd
edit: manuscript

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of multiple factors. For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems than small ones, whereas FCL was more sensitive to size in low-productivity systems. Such context-dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of FCL [@takimotoEffectsProductivityDisturbance2012; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure assumes a simple linear food chain with a maximum length of three (**Figure 1D**). However, FCL decreases proportionally with increasing omnivory as the top predator eats lower in the food web [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is too strong, leading to the loss of species and shorter food chains [@woottonOmnivoryStabilityFreshwater2017]. Conversely, weak omnivory stabilizes the food web and increases the persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers and the potential for interactive effects, a unified framework of multiple FCL drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

While recent research efforts of food web ecologists have substantially advanced our understanding of food chain drivers, most of the previous work has been conducted in simple, two-dimensional systems, such as lakes [@postEcosystemSizeDetermines2000; @doiResourceAvailabilityEcosystem2009] and oceanic islands [@takimotoEcosystemSizeNot2008]. However, many natural habitats have a high degree of spatial complexity that cannot be represented simply by ecosystem size. Branching ecosystems, for example, are ubiquitous yet overlooked landscape structures (e.g., rivers, trees, caves, mountain ranges) [@grantLivingBranchesPopulation2007; @altermattDiversityRiverineMetacommunities2013]. While we know that food web interactions are spatially influenced [@calcagnoConstraintsFoodChain2011; @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012; @teruiSpatialDisturbanceSynchrony2019], the theory to account for it has lagged, particularly in branching habitat architectures. Here, we theoretically explore drivers of FCL in rivers, an excellent example of branching ecosystems.

@@

There were three consistent patterns in the relationship between FCL and ecosystem geometry (size and complexity). First, the effects of ecosystem size and complexity were modulated by disturbance and omnivory. Second, resource availability and dispersal had limited influences on the qualitative relationship between FCL and ecosystem geometry, although they had important direct effects on the FCL. Below we describe these results in detail.

@@

Ecosystem geometry interacted with disturbance to dictate FCL in branching ecosystems. With no disturbance, both geometric features (size and complexity) had limited influence on FCL. However, as disturbance frequency increases, the relationships between FCL and ecosystem geometry became positive. This interactive effect emerged because the sensitivity of FCL to disturbance depended on ecosystem geometry. Although increased disturbance frequency reduced overall FCL, smaller and/or simpler ecosystems exhibited the greater vulnerability. Therefore, disturbance interact with ecosystem geometry, producing the interdependency of these factors in determining FCL. The interactive effects, however, were not observed with low disturbance intensity (proportion of individuals removed per disturbance event); this suggests that disturbance intensity must be strong enough to be influential.

The exception is the "chain" scenario (i.e., no omnivory) with no disturbance. In this scenario, ecosystem size had a clear positive effect on FCL while ecosystem complexity had a negative influence. These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increases the total carrying capacity whereas ecosystem complexity decreases it. However, these patterns characteristic to the "chain" scenario diminished when considering some levels of omnivory because food webs with omnivory were less reliant on the carrying capacity of the basal species.

@@

Another interesting result of our simulation is the interaction with omnivory. In particular, when the chain motif was combined with no-disturbance, ecosystem size and complexity had contrasting influences on FCL. It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity; therefore, the total carrying capacity within a network dictates FCL. The effect of ecosystem size is understandable because it simply increases the total carrying capacity. In contrast, ecosystem complexity, here assessed as branching probability, decreases the total carrying capacity as more branched networks have disproportionate numbers of headwater habitats with small local carrying capacities. However, the dependence of FCL on local carrying capacity diminished in weak and strong omnivory scenarios, explaining the lack of positive (ecosystem size) or negative FCL-geomtery relationship (ecosystem complexity) in these motifs with no disturbance.

@@

The effect of productivity was context dependent, supporting previous theoretical predictions. While productivity clearly increased FCL in the chain motif, this effect was vague (strong omnivory) or even reversed (weak omnivory). High productivity effectively increases the abundance of the basal species, stimulating the prey switching of the intraguild predator from the intraguild prey to the basal species. As such, increased productivity does not necessarily increase FCL. Our results suggest that this prediction from non-spatial models can be applied to spatially complex ecosystems.

Ecosystem size and complexity led to the distinct patterns of state transitions. The state transitions with increased ecosystem size are likely associated with increased immigration. It is clear that the total number of dispersing propagules increases with increasing ecosystem size, providing opportunities to maintain longer food chains. The exception was the strong omnivory motif with rare disturbance, in which the insertion mechanism operated. This is likely because rare disturbance inhibited the exclusion of the intraguild prey by suppressing the abundance of the intraguild predator. Similarly, increased immigration may be generally responsible for the state transitions with increasing ecosystem complexity. Habitat patches are closer in more branched networks as it compresses the network configuration. However, unlike ecosystem size, ecosystem complexity does not increase the total number of dispersing propagules - rather, it enhances connectivity among habitat patches. This difference may explain the distinct patterns of state transitions along the gradients of ecosystem size and complexity.

>>>>

More recently, food web ecologists have begun to recognize that FCL rarely responds to a single driver, but instead is determined by interactions of multiple factors. For example, @wardMechanisticTheoryAquatic2017 showed that FCL was more responsive to increasing productivity in large ecosystems than small ones, whereas FCL was more sensitive to size in low-productivity systems. Such context-dependent responses to extrinsic factors seem to be tightly coupled with the arrangement of trophic links within a food web. For example, omnivory (feeding at multiple trophic levels) is a strong intrinsic driver of FCL [@takimotoEffectsProductivityDisturbance2012; @postProximateStructuralMechanisms2007; @woottonOmnivoryStabilityFreshwater2017]. When omnivory is absent in three-species communities, the trophic structure or "motif" assumes a simple linear food chain with a maximum trophic position of three (**Figure 1D**). However, FCL decreases with increasing omnivory as the top predator eats lower in the food web [@wardMechanisticTheoryAquatic2017; @woottonOmnivoryStabilityFreshwater2017]. Theoretical work predicts that food webs are destabilized when omnivory is too strong, leading to the loss of species and shorter food chains [@woottonOmnivoryStabilityFreshwater2017]. Conversely, weak omnivory stabilizes the food web and increases the persistence of species within a community, particularly at intermediate levels of productivity and high disturbance levels [@woottonOmnivoryStabilityFreshwater2017; @woottonManyWeakInteractions2016]. Due to the context-dependency of FCL to multiple drivers and the potential for interactive effects, a unified framework of multiple FCL drivers is needed to better understand when and where FCL will respond to these extrinsic and intrinsic factors.

While recent research efforts of food web ecologists have substantially advanced our understanding of food chain drivers, theoretical conceptualization has implicitly assumed two-dimensional systems, which may be a fair approximation of lakes [@postEcosystemSizeDetermines2000; @doiResourceAvailabilityEcosystem2009] and oceanic islands [@takimotoEcosystemSizeNot2008]. However, many natural habitats have a high degree of spatial complexity that cannot be represented simply by ecosystem size. Branching ecosystems, for example, are ubiquitous yet overlooked landscape structures (e.g., rivers, trees, caves, mountain ranges) [@grantLivingBranchesPopulation2007; @altermattDiversityRiverineMetacommunities2013]. While we know that food web interactions are spatially influenced [@calcagnoConstraintsFoodChain2011; @pillaiMetacommunityTheoryExplains2011; @takimotoEffectsProductivityDisturbance2012; @teruiSpatialDisturbanceSynchrony2019], the theory to account for it has lagged, particularly in branching habitat architectures. Here, we theoretically explore drivers of FCL in rivers, an excellent example of branching ecosystems.

@@

There were two consistent patterns in the relationship between FCL and ecosystem geometry (size and complexity). First, the effects of ecosystem size and complexity were modulated by disturbance and omnivory. Second, resource availability and dispersal had limited influences on the qualitative relationship between FCL and ecosystem geometry, although they had important direct effects on the FCL. Below we describe these results in detail.

@@

Ecosystem geometry interacted with disturbance to dictate FCL in branching ecosystems (**Figures 2 and 3**). With no disturbance, both geometric features (size and complexity) had limited influence on FCL. However, as disturbance frequency increases, the relationships between FCL and ecosystem geometry became positive. This interactive effect emerged because the sensitivity of FCL to disturbance depended on ecosystem geometry. Although increased disturbance frequency reduced overall FCL, smaller and/or simpler ecosystems exhibited the greater vulnerability (**Figures 2 and 3**). Therefore, disturbance interacted with ecosystem geometry to produce the interdependency of these factors in determining FCL. However, the maximum levels of FCL was strongly regulated by disturbance frequency, highlighting the important direct effect of disturbance. It is important to note that the interactive effects were not observed with low disturbance intensity (proportion of individuals removed per disturbance event); this suggests that disturbance intensity must be strong enough to be influential.

The exception is the "chain" scenario (i.e., no omnivory) with no disturbance. In this scenario, ecosystem size had a clear positive effect on FCL (**Figure 2**) while ecosystem complexity had a negative influence (**Figure 3**). These opposing influences of ecosystem size and complexity were associated with their contrasting relationships with the total carrying capacity (i.e., carrying capacity summed across all patches): ecosystem size increases the total carrying capacity whereas ecosystem complexity decreases it (**Figure XXX**). However, these patterns characteristic to the "chain" scenario diminished when considering some levels of omnivory because FCLs with omnivory were less reliant on the carrying capacity of the basal species.

@@

Another interesting result of our simulation is the interaction with omnivory. In particular, when the chain motif was combined with no-disturbance, ecosystem size and complexity had contrasting influences on FCL. It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity; therefore, the total carrying capacity within a network dictates FCL. This effect of ecosystem size is understandable because it simply increases the total carrying capacity. In contrast, ecosystem complexity, here assessed as branching probability, decreased the total carrying capacity as more branched networks have disproportionate numbers of headwater habitats with small local carrying capacities. Thus, ecosystem complexity had an opposite effect on FCL with the chain motif. However, the dependence of FCL on local carrying capacity diminished in weak and strong omnivory scenarios. This fact explains the lack of positive (ecosystem size) or negative FCL-geomtery relationship (ecosystem complexity) in these motifs with no disturbance.

@@

The effect of productivity was context-dependent, supporting previous theoretical predictions [@postProximateStructuralMechanisms2007; @kondohFoodchainLengthAdaptive2009; @wardMechanisticTheoryAquatic2017]. While productivity clearly increased FCL in the chain motif, this effect was vague (strong omnivory) or even reversed (weak omnivory) in omnivory motifs. With omnivory, high productivity effectively increases the abundance of the basal species, stimulating the prey switching of the intraguild predator from the intraguild prey to the basal species. As such, increased productivity does not necessarily increase FCL. Our results suggest that this prediction from non-spatial models can be applied to spatially complex ecosystems.

Ecosystem size and complexity led to the distinct patterns of state transitions. The state transitions with increased ecosystem size are likely associated with increased immigration. The total number of dispersing propagules increases with increasing ecosystem size, providing opportunities to maintain longer food chains in connected patches. The exception was the strong omnivory motif with rare disturbance, in which the insertion mechanism operated. This is likely because rare disturbance inhibited the exclusion of the intraguild prey by suppressing the abundance of the intraguild predator. Similarly, increased immigration may be generally responsible for the state transitions with increasing ecosystem complexity. Habitat patches are closer in more branched networks as branching compresses the network configuration. However, unlike ecosystem size, ecosystem complexity does not increase the total number of dispersing propagules - rather, it enhances connectivity among habitat patches. This difference may explain the distinct patterns of state transitions along the gradients of ecosystem size and complexity.
